---
title: "Final Project"
output: html_document
date: "2025-05-17"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Setting up

```{r cars}
library(tidyverse)
library(ggplot2)
library(readxl)
library(dplyr)
library(readr)
library(lubridate)

# Load the Dataset
salesMonthly <- read_csv("~/Desktop/Syracuse/IST 687/Pharma Sales (Kaggle)/salesmonthly.csv")
salesWeekly <- read_csv("~/Desktop/Syracuse/IST 687/Pharma Sales (Kaggle)/salesweekly.csv")
salesDaily <- read_csv("~/Desktop/Syracuse/IST 687/Pharma Sales (Kaggle)/salesdaily.csv")
salesHourly <- read_csv("~/Desktop/Syracuse/IST 687/Pharma Sales (Kaggle)/saleshourly.csv")

# Quick Peek(s)
View(salesMonthly)
View(salesWeekly)
View(salesDaily)
View(salesHourly)
```

## Initial Assumption(s)

```{r}
# Codes:
  #M01AB - Anti-inflammatory and antirheumatic products, non-steroids, Acetic acid derivatives and related substances
  #M01AE - Anti-inflammatory and antirheumatic products, non-steroids, Propionic acid derivatives
  #N02BA - Other analgesics and antipyretics, Salicylic acid and derivatives
  #N02BE/B - Other analgesics and antipyretics, Pyrazolones and Anilides
  #N05B - Psycholeptics drugs, Anxiolytic drugs
  #N05C - Psycholeptics drugs, Hypnotics and sedatives drugs
  #R03 - Drugs for obstructive airway diseases
  #R06 - Antihistamines for systemic use Sales data are resampled to the hourly, daily, weekly and monthly periods. Data is already pre-processed, where processing included outlier detection and treatment and missing data imputation.

# Predictive Trends to Watch For
  #R03 - peak in Spring/Summer
    #R06 - peak in Fall/Winter
  #N02BA, N02BE - painkillers hold stable
  #M01AB, M01AE - anti-inflammatory, small increase during peak Flu season (Fall, Winter)
  #N05B, N05C - Psycholeptics - stable with peaks during Winter, trending upwards over time
```

## Initial Prep

```{r pressure, echo=FALSE}
# Date Formatting
salesMonthly <- salesMonthly %>%
  mutate(
    datum = as.Date(datum),
    Month = month(datum, label = TRUE),
    Year = year(datum),
    Season = case_when(
      Month %in% c("Dec", "Jan", "Feb") ~ "Winter",
      Month %in% c("Mar", "Apr", "May") ~ "Spring",
      Month %in% c("Jun", "Jul", "Aug") ~ "Summer",
      Month %in% c("Sep", "Oct", "Nov") ~ "Fall"
    )
  )
```


## Seasonal Trends I - RO3, R06 - Respiratory Drugs

```{r}
# Plot Pivot
seasonal_long <- salesMonthly %>%
  select(datum, Season, R03, R06) %>%
  pivot_longer(cols = c(R03, R06), names_to = "Drug", values_to = "Sales")

# Plotting Averages
seasonal_long %>%
  group_by(Season, Drug) %>%
  summarise(Average_Sales = mean(Sales, na.rm = TRUE), .groups = "drop") %>%
  ggplot(aes(x = Season, y = Average_Sales, fill = Drug)) +
  geom_col(position = "dodge") +
  labs(title = "Average Seasonal Sales for Respiratory Drugs (R03, R06)",
       x = "Season", y = "Average Monthly Sales") +
  theme_minimal()

# R06 - clear trends in Spring/Summer (likely from "allergy season")
# R03 - clear spikes in "colder months" where respiratory infections and asthma triggering weather is more likely
```

## Seasonal Trends II - M01AB, M01AE - NSAIDs (Pain Management Drugs)

```{r}
# Plot Pivot
seasonal_long <- salesMonthly %>%
  select(datum, Season, M01AB, M01AE) %>%
  pivot_longer(cols = c(M01AB, M01AE), names_to = "Drug", values_to = "Sales")

# Plotting Averages
seasonal_long %>%
  group_by(Season, Drug) %>%
  summarise(Average_Sales = mean(Sales, na.rm = TRUE), .groups = "drop") %>%
  ggplot(aes(x = Season, y = Average_Sales, fill = Drug)) +
  geom_col(position = "dodge") +
  labs(title = "Average Seasonal Sales for NSAIDs (M01AB, M01AE)",
       x = "Season", y = "Average Monthly Sales") +
  theme_minimal()

# M01AB - stable with a small peak in Summer, drop in Winter
# M01AE - mild increase in Winter (Flu season)
```

## Seasonal Trends III - N02BA, N02BE - Salicylic acid derivatives, Anilides & Pyrazolones  (Fever/Pain Management Drugs)

```{r}
# Plot Pivot
seasonal_long <- salesMonthly %>%
  select(datum, Season, N02BA, N02BE) %>%
  pivot_longer(cols = c(N02BA, N02BE), names_to = "Drug", values_to = "Sales")

# Plotting Averages
seasonal_long %>%
  group_by(Season, Drug) %>%
  summarise(Average_Sales = mean(Sales, na.rm = TRUE), .groups = "drop") %>%
  ggplot(aes(x = Season, y = Average_Sales, fill = Drug)) +
  geom_col(position = "dodge") +
  labs(title = "Average Seasonal Sales for SA, Anilides & Pyrazolones (N02BA, N02BE)",
       x = "Season", y = "Average Monthly Sales") +
  theme_minimal()

# Notable gap between purchase trends in amount compared between the two 
# N02BA - low purchase trends throughout the year
# N02BE - highest purchasing trends (of any drugs in this dataset) throughout the year with peaks in the Fall/Winter (Flu Season)
```

## Seasonal Trends IV - N05B, N05C - Psycholeptics – Anxiolytics, Hypnotics & Sedatives

```{r}
# Plot Pivot
seasonal_long <- salesMonthly %>%
  select(datum, Season, N05B, N05C) %>%
  pivot_longer(cols = c(N05B, N05C), names_to = "Drug", values_to = "Sales")

# Plotting Averages
seasonal_long %>%
  group_by(Season, Drug) %>%
  summarise(Average_Sales = mean(Sales, na.rm = TRUE), .groups = "drop") %>%
  ggplot(aes(x = Season, y = Average_Sales, fill = Drug)) +
  geom_col(position = "dodge") +
  labs(title = "Average Seasonal Sales for Psycholeptics – Anxiolytics, Hypnotics & Sedatives (N05B, N05C)",
       x = "Season", y = "Average Monthly Sales") +
  theme_minimal()

# N05B - much higher steady trends with peak during fall - possibly related to holidays?
# N05C - low steady trends throughout the year with a "slight" peak in winter
```
```{r}
names(df)
```

```{r}
#weekly regression = sleep aids
lm_sleep <- lm(Sleep_Aids ~ factor(`Weekday.Name`) + factor(Month), data = df)
summary(lm_sleep)
```


```{r}
#weekly regression = anxiety meds
lm_anxiety <- lm(N05B ~ factor(`Weekday.Name`) + factor(Month), data = df)
summary(lm_anxiety)
```

```{r}
#yearly regression = anxiety meds
lm_anxiety2 <- lm(N05B ~ factor(`Month`) + factor(Year), data = df)
summary(lm_anxiety2)
```


```{r}
#Histogram, Summary, Table Analysis of NSAIDs
hist(df$M01AB)

sum(is.null(df$M01AB))

table(df$M01AB)
```

```{r}
#Sales Frequency Trends for Allergy Medications
yt <- ts(df$R06, frequency = 365, start = c(2014,1,1))
plot(yt)
```

```{r}
#HoltWinters Forecast for Allergy Medications
ts_forecast <- HoltWinters(yt, alpha=TRUE, 
                            beta=FALSE, 
                            gamma=TRUE)

ts_forecast
plot(ts_forecast)
ts_forecast1 <- forecast(ts_forecast, h=100)
plot(ts_forecast1)
```
```{r}
#Arima Analysis of Allergy Medications
y_diff <- diff(yt)
plot(y_diff)
acf(y_diff)
```

```{r}
#Arima Analysis of Allergy Medications II
y_ma <- arima(y_diff, order = c(0,0,1))
plot(yt)
lines(yt-y_ma$residuals, col="red")
```

```{r}
#Forecast Trend Analysis for Allergy Medications
d_n = forecast(y_ma, h = 1)
y_next = d_n$mean + yt[length(yt)]
y_next = as.numeric(y_next)
y_next
```

```{r}
#Arima Plot Comparison - Allergy Medications
y_ma = arima(y_diff, order = c(0,0,1))
y_ar = arima(y_diff, order = c(1,0,0))
# plot the fitted series
plot(yt)
lines(yt-y_ma$residuals, col = "red")
lines(yt-y_ar$residuals, col = "blue")
```

```{r}
d_n = forecast(y_ar, h = 1)
y_next = d_n$mean + yt[length(yt)]
y_next = as.numeric(y_next)
y_next
## [1] 2.916789
```

```{r}
ytM01AB <- ts(df$M01AB, frequency = 365, start = c(2014,1,1))

ytM01AE <- ts(df$M01AE, frequency = 365, start = c(2014,1,1))

ytN02BA <- ts(df$N02BA, frequency = 365, start = c(2014,1,1))

ytN02BE <- ts(df$N02BE, frequency = 365, start = c(2014,1,1))

ytN05B <- ts(df$N05B, frequency = 365, start = c(2014,1,1))

ytN05C <- ts(df$Sleep_Aids, frequency = 365, start = c(2014,1,1))

ytR03 <- ts(df$R03, frequency = 365, start = c(2014,1,1))

ytR06 <- ts(df$R06, frequency = 365, start = c(2014,1,1))
```

```{r}
plot(ytM01AB)
plot(ytM01AE)
plot(ytN02BA)
plot(ytN02BE)
plot(ytN05B)
plot(ytN05C)
plot(ytR03)
plot(ytR06)
```

```{r}
sapply(df,summary)
```
```{r}
for (col in names(df)[2:9]) {
  p <- ggplot(df, aes_string(x = col)) +
    geom_histogram() +
    labs(title = paste(col),
         x = col)
  
  print(p)
}
```

```{r}
library(tidyverse)
library(readxl)
library(forecast)
salesdaily <- read.csv("~/Desktop/Syracuse/IST 687/Pharma Sales (Kaggle)/salesdaily.csv")
df <- salesdaily
```

```{r}
# Separating into two training and validation
dfTrain <- df %>% slice(1:1825)
dfValidate <- df %>% slice(1826:nrow(df))


yt <- ts(dfTrain$R06, frequency = 365, start = c(2014,1,1))
plot(yt)
```

```{r}
#HoltWinters for Allergy Medications - Comparison before and after trainig the model
ts_forecast <- HoltWinters(yt, alpha=TRUE, 
                            beta=FALSE, 
                            gamma=TRUE)

ts_actual <- ts(dfValidate$R06, frequency = 365, start = c(2019,1,1))

  
ts_forecast1 <- forecast(ts_forecast, h=281)


plot(ts_forecast1);lines(ts_actual, col = 2)

difference <- ts_forecast1$mean - ts_actual
mean(difference)
```

```{r}
file_path <- "~/Desktop/Syracuse/IST 687/Pharma Sales (Kaggle)/salesdaily.csv"
df<-read_csv(file_path)
library(readr)
library(tidyverse)
df <- df %>%
  rename(
    Acetic_NSAIDs       = M01AB,
    Propionic_NSAIDs    = M01AE,
    Aspirin_Group       = N02BA,
    Paracetamol_Group   = N02BE,
    Anxiety_Drugs       = N05B,
    Sleep_Aids          = N05C,
    Asthma_Drugs        = R03,
    Allergy_Meds        = R06
  )

df$datum <- lubridate::mdy(df$datum)
glimpse(df)
```
```{r}
#2. Mental Health Medication Usage Over Time

df_long2 <- df %>%
  select(datum, Anxiety_Drugs, Sleep_Aids) %>%
  pivot_longer(-datum, names_to = "Drug_Type", values_to = "Sales")

ggplot(df_long2, aes(x = datum, y = Sales, color = Drug_Type, group = Drug_Type)) +
  geom_line(alpha = 0.8) +
  scale_x_date(date_breaks = "6 months", date_labels = "%b %Y") +
  labs(title = "Daily Sales: Mental Health Medications", x = "Date", y = "Sales") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
```{r}
library(tidyverse)
library(lubridate)

df_long <- df %>%
  select(datum, Acetic_NSAIDs, Propionic_NSAIDs, Aspirin_Group, Paracetamol_Group) %>%
  pivot_longer(-datum, names_to = "Drug_Type", values_to = "Sales")

ggplot(df_long, aes(x = datum, y = Sales, color = Drug_Type, group = Drug_Type)) +
  geom_line(alpha = 0.8) +
  scale_x_date(date_breaks = "6 months", date_labels = "%b %Y") +
  labs(title = "Trends of Pain Relief Drugs", x = "Date", y = "Sales") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
1. Paracetamol (cyan line) dominates sales:
It is consistently the highest-selling drug group over the entire period.

Sales show seasonal spikes (e.g., visible in early 2015, 2016, 2017, and late 2018), likely during cold/flu seasons or other viral outbreaks.

One large spike in early 2017 likely reflects a significant demand event — possibly due to a widespread illness or policy change.

2. NSAIDs (Acetic & Propionic) and Aspirin_Group have:
Lower but steady usage over time.

No major upward or downward trend — usage remains fairly stable.

These lines hover around a baseline, showing less seasonal variation than paracetamol.

3. No major structural shifts over time:

No medication shows a dramatic upward or downward long-term trend.

Interpretation / Insights:
Paracetamol sales show short-term bursts, while the others remain flat with minor fluctuations.

Paracetamol appears to be the first-line pain/fever remedy for the population — likely due to OTC availability and broad usage.

Spikes in Paracetamol_Group align with expected flu seasons (winter to early spring).

NSAIDs and Aspirin may be more commonly used for chronic or inflammatory conditions, leading to steady, non-seasonal patterns.


```{r}

library(tidyverse)
library(lubridate)
library(readr)
df <- read_csv("~/Desktop/Syracuse/IST 687/Pharma Sales (Kaggle)/salesdaily.csv", 
               col_types = cols(datum = col_character()))
df$datum <- mdy(df$datum)
summary(df$datum)

df$Month <- month(df$datum, label = TRUE, abbr = TRUE)

df_long3 <- df %>%
  select(Month, R03, R06) %>%
  pivot_longer(cols = c(R03, R06), 
               names_to = "Drug_Type", values_to = "Sales")

ggplot(df_long3, aes(x = Month, y = Sales, fill = Drug_Type)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.7) +
  labs(title = "Seasonal Sales Trends: Respiratory and Allergy Meds",
       x = "Month", y = "Sales") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```
Asthma_Drugs:

Higher variability and sales overall.

Peaks in winter and early spring (Dec–Mar), with December showing the highest sales.

Suggests seasonal spikes potentially tied to cold/flu season, winter triggers, or increased respiratory infections.

Allergy_Meds:

More stable, lower median sales throughout the year.

Slight uptick in spring (Mar–May) — potentially aligning with pollen/allergy season.

Sales remain relatively flat compared to asthma drugs.

Interpretation
Asthma medication usage is strongly seasonal, with clear increases during colder months. This could be linked to:

Respiratory infections

Cold-induced asthma exacerbation

Increased indoor allergens (e.g., dust, mold)

Allergy medications show minor seasonal influence, possibly indicating:

More consistent year-round use (chronic allergies)

Mild seasonal variation with a small rise in spring due to environmental allergens like pollen.

```{r}
#Cluster Map
df_cluster_input <- df %>%
  select(M01AB, M01AE, N02BA, N02BE,
         N05B, N05C, R03, R06) %>%
  scale() %>%
  as_tibble()

set.seed(123)
k_result <- kmeans(df_cluster_input, centers = 4, nstart = 25)
df$cluster <- as.factor(k_result$cluster)

df$total_sales <- rowSums(df[, c("M01AB", "M01AE", "N02BA", "N02BE",
         "N05B", "N05C", "R03", "R06")])

df$Cluster_Label <- recode(df$cluster,
  "1" = "1 - Moderate Sales",
  "2" = "2 - Low Sales",
  "3" = "3 - High Sales",
  "4" = "4 - Very Low Sales"
)

ggplot(df, aes(x = datum, y = total_sales, color = Cluster_Label)) +
  geom_point(alpha = 0.6) +
  labs(
    title = "Clustered Days Based on Drug Use Profiles",
    x = "Date", y = "Total Drug Sales", color = "Cluster"
  ) +
  scale_x_date(date_breaks = "1 year", date_labels = "%Y") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
Cluster 3 (cyan/light blue):
Represents high-sales days, likely corresponding to flu season spikes (e.g., winter months like Jan–Mar).

These days could be driven by Paracetamol_Group or Asthma_Drugs increases.

Interpretation: Likely high demand due to respiratory infections, flu outbreaks, or seasonal illness peaks.

Cluster 1 (light red):
Shows a moderate-high sales pattern, but not as extreme as Cluster 3.

Possibly includes weekends or small public health events.

Interpretation: Elevated baseline usage — maybe moderate seasonal triggers or OTC activity.

Cluster 2 (light green):
Appears around the median sales range with relatively stable values.

Interpretation: Normal, steady-state drug usage. Possibly routine or chronic-use medication periods.

Cluster 4 (purple):
Clearly the lowest total sales cluster.

Possibly off-season periods or days with low retail activity (e.g., holidays, reporting gaps, or weekends).

Interpretation: Low-demand days — not much medication being purchased or dispensed.



```{r}
#Seasonal Sales Numbers = All Medications
df_summary <- df %>%
  group_by(cluster) %>%
  summarise(across(
    c(M01AB, M01AE, N02BA, N02BE,
         N05B, N05C, R03, R06),
    mean,
    .names = "avg_{.col}"
  ))

print(df_summary)
```
Cluster: High Flu Season Days
Paracetamol_Group has the highest average → Suggests cold/flu symptom treatment.

Asthma_Drugs are also elevated → Flu can worsen asthma.

Interpretation: Likely January–March. Respiratory illness peaks. OTC cold remedies dominate.

Cluster: Mental Health Days

Anxiety_Drugs and Sleep_Aids are the highest here.

Pain meds like NSAIDs or Paracetamol are lower.

Interpretation: Possibly stress-related spikes (e.g., late-year holidays or post-summer adjustment periods).

Cluster: Baseline/Low Sales

Most drug groups have their lowest averages.

Interpretation: Quiet months like summer (June–August), with minimal illness and fewer prescriptions.

Cluster: Steady Pain Relief

Aspirin, NSAIDs, and Paracetamol show consistent, mid-level averages.

Interpretation: Reflects days where pain and chronic condition management dominate.


```{r}
#Heat Maps = All Medications
df_summary <- df %>%
  group_by(Cluster_Label) %>%
  summarise(across(
    c(M01AB, M01AE, N02BA, N02BE,
         N05B, N05C, R03, R06),
    mean,
    .names = "avg_{.col}"
  ))

df_heatmap <- df_summary %>%
  pivot_longer(cols = starts_with("avg_"),
               names_to = "Drug_Type",
               values_to = "Avg_Sales") %>%
  mutate(Drug_Type = str_remove(Drug_Type, "avg_"))

ggplot(df_heatmap, aes(x = Drug_Type, y = Cluster_Label, fill = Avg_Sales)) +
  geom_tile(color = "white") +
  scale_fill_gradient(low = "lightblue", high = "darkred") +
  labs(
    title = "Heatmap of Average Drug Sales by Cluster",
    x = "Drug Type", y = "Cluster", fill = "Avg Sales"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```
The heatmap reveals distinct usage patterns across drug categories segmented by sales intensity clusters:

- **Cluster 3 – High Sales**  
  This cluster shows the highest average drug sales, with a strong concentration in **Paracetamol_Group**. This pattern suggests flu or cold season spikes, likely during winter months. **Asthma_Drugs** and **Anxiety_Drugs** also show moderate elevation, possibly due to increased respiratory stress during colder seasons.

- **Cluster 1 – Moderate Sales**  
  Drug usage is more balanced in this cluster. **Anxiety_Drugs** stand out with higher average values compared to other clusters, potentially indicating general stress-related or chronic mental health demand. **Paracetamol_Group** also remains elevated, though not as intense as in Cluster 3.

- **Cluster 2 – Low Sales**  
  Sales levels drop for all drugs, though **Paracetamol_Group** maintains a modest presence. This cluster likely reflects regular days without major illness surges or seasonal effects.

- **Cluster 4 – Very Low Sales**  
  This cluster exhibits the lowest drug sales across all categories, suggesting minimal demand. These may correspond to weekends, holidays, or summer periods when illness and medication use typically decline.

**Key Insight:**  
Paracetamol dominates high-sales clusters, supporting its role as a first-line response for common illnesses. Anxiety and sleep medications peak during moderate usage periods, reflecting steady, non-seasonal demand. NSAID groups show consistent but low variation across clusters, implying their use is less tied to seasonal illness patterns.

```{r}
psych1 <- lm(data = df, N05B ~ N02BA + R06)
 
summary(psych1)
```

